The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with \textit{Gaia} Data and Beyond

### 2019-05-13 paper/ms.tex
Swapped Melissa's affil
revisions to abstract, intro (Sect 1)

\author[0000-0003-0064-4060]{Mordecai-Mark Mac Low}
\affil{\cca}
\affil{\amnh}

@@

Orbital properties of stars, computed from their six-dimensional phase space
measurements and an assumed Galactic potential, are used to understand the
structure and evolution of the Galaxy. Stellar actions, computed from orbits,
have the attractive quality of being invariant and thus good quantitative labels
of a star's orbit. A systematic bias in the phase space position induces a
phase-dependence in the actions, which we interpret as a systematic error
distinct from typical observational errors and Galactic potential uncertainties.
An offset in the vertical position of $\sim15\,\pc$ ($\sim120\,\pc$) for a thin
(thick) disk orbit is sufficient to induce a $25\%$ systematic error in the
vertical action $J_z$. The induced error distribution is non-Gaussian and
bimodal, with neither of the modes peaking on the null value. Furthermore, we
show that the local midplane varies by $\sim100\,\pc$ at the solar circle in
Milky Way-mass cosmological zoom-in simulations from the FIRE project. From
observations that the mean vertical velocity of the Milky Way disk varies by
$\sim5\textup{--}10\,\kms$ with radius, we estimate that the true midplane
variations are of order $\sim60\textup{--}170\,\pc$, with similar offsets seen
in recent three-dimensional dust maps. Thus, current state-of-the-art action
calculations, which assume the global and local midplanes are identical, are
likely to include a systematic vertical offset depending on the volume
considered. Variation in the local standard of rest induces similar issues. The
variation of the midplane must be taken into account when performing dynamical
analysis across the large regions of the disk accessible to \textit{Gaia} and
future missions.

@@

The assumption of a global, axisymmetric Galactocentric coordinate system
\citep{2008gady.book.....B} underlies much of the quantitative analysis of the
mechanisms that give rise to these signautres. In order to do construct such a
system, the Sun's relative position must be defined and measured both precisely
and accurately. This involves determining the angular position of and distance
to the Galactic center, the orientation of the Galactic plane, the height of the
Sun above the Galactic midplane, and the local standard of rest (LSR). We review
and discuss the observational efforts to measure these parameters in
Section~\ref{ssec:coord_off}.

@@

The true Galactic potential is not strictly axisymmetric. This has been known
for some time \citep[e.g.,][]{1957AJ.....62...93K,2009MNRAS.396L..56M,
2012ApJ...750L..41W}, but with the vast improvement in the quality of
phase-space measurements due to \textit{Gaia} this assumption is increasingly
inadequate \citep[e.g.,][]{2018Natur.561..360A, 2019MNRAS.485.3134L}. Even if
this assumption were close enough for many purposes, the parameters used in
axisymmetric models of the Galactic potential may be inadequately constrained by
current observations.

The main point of this paper is as follows. Due to relatively small-scale
variations in the density of gas and stars, the local midplane should vary as a
function of radius and azimuth. For stars relatively far from the Sun, their
local midplane will differ from our local midplane extrapolated onto their
position. Converting positions and velocities of more distant stars from a
heliocentric to a Galactocentric coordinate system thus introduces a systematic
bias in the $z$ coordinate. We show that this bias induces non-Gaussian errors
in the actions computed for these stars. The further from the solar neighborhood
the target star is, the more likely the mismatch will result in large systematic
uncertainty, especially in the vertical action $J_z$. A similar argument applies
to any remaining uncertainty in measurements of the Galactocentric radius of the
Sun and to variations in the LSR.

>>>>

\author[0000-0003-0064-4060]{Mordecai-Mark Mac Low}
\affil{\amnh}
\affil{\cca}

@@

Orbital properties of stars, computed from their six-dimensional phase space
measurements and an assumed Galactic potential, are used to understand the
structure and evolution of the Galaxy. Stellar actions, computed from orbits,
have the attractive quality of being invariant and thus good quantitative labels
of a star's orbit. A systematic bias in the phase space position
induces 
%mm a phase-dependence [to better distinguish the two different
%meanings of phase in this sentence!]
   an orbital phase dependence
in the actions, which we interpret as a systematic error
distinct from typical observational errors and Galactic potential uncertainties.
An offset in the vertical position of 
%mm
     the Galactic midplane of
$\sim15\,\pc$ 
%mm [this formulation isn't worth the extra difficulty reading]
%($\sim120\,\pc$) for a thin (thick) disk 
  for a thin disk orbit or $\sim 120\,\pc$ for a thick disk
orbit is sufficient to induce a $25\%$ systematic error in the
vertical action $J_z$. The induced error distribution is non-Gaussian and
bimodal, with neither of the modes peaking on the null value. Furthermore, we
show that the local midplane varies by $\sim100\,\pc$ at the solar circle in
Milky Way-mass cosmological zoom-in simulations from the FIRE project. From
observations that the mean vertical velocity of the Milky Way disk varies by
$\sim5\textup{--}10\,\kms$ with radius, we estimate that the true midplane
variations are of order $\sim60\textup{--}170\,\pc$, with similar offsets seen
in recent three-dimensional dust maps. Thus, current state-of-the-art action
calculations, which assume the global and local midplanes are identical, are
likely to include a systematic vertical offset depending on the volume
considered. Variation in the local standard of rest induces similar issues. The
variation of the midplane must be taken into account when performing dynamical
analysis across the large regions of the disk accessible to \textit{Gaia} and
future missions.

@@

The assumption of a global, axisymmetric Galactocentric coordinate system
\citep{2008gady.book.....B} underlies much of the quantitative analysis of the
mechanisms that give rise to these 
%mm signautres.
    signatures. 
In order to 
%mm do 
construct such a
system, the Sun's relative position must be defined and measured both precisely
and accurately. This involves determining the angular position of and distance
to the Galactic center, the orientation of the Galactic 
%mm plane, the height of the Sun above the Galactic midplane, 
   midplane, the distance of the Sun from it,
and the local standard of rest (LSR). We review
and discuss the observational efforts to measure these parameters in
Section~\ref{ssec:coord_off}.

@@

The true Galactic potential is not strictly axisymmetric. This has been known
for some time \citep[e.g.,][]{1957AJ.....62...93K,2009MNRAS.396L..56M,
2012ApJ...750L..41W}, but with the vast improvement in the quality of
phase-space measurements due to \textit{Gaia} 
%mm this assumption 
     the assumption of axisymmetry
is increasingly inadequate \citep[e.g.,][]{2018Natur.561..360A, 2019MNRAS.485.3134L}. Even if
this assumption were close enough for many purposes, the parameters used in
axisymmetric models of the Galactic potential may be inadequately constrained by
current observations.

%mm The main point of this paper is as follows. Due to relatively
%small-scale variations in the density of gas and stars, the local
%midplane should vary as a
      Small-scale variations in the density of gas and stars cause the
      local midplane position to vary as a
function of radius and azimuth. 
%mm For stars relatively far from the Sun, their local midplane will
%differ 
    Stars far from the Sun have a local midplane that differs
from our local midplane extrapolated onto their position. Converting
positions and velocities of more distant stars from a 
heliocentric to a Galactocentric coordinate system thus introduces a systematic
bias in the $z$ coordinate. We show that this bias induces non-Gaussian errors
in the actions computed for these stars. The further from the solar neighborhood
the target star is, the more likely the mismatch will result in large systematic
uncertainty, especially in the vertical action $J_z$. A similar argument applies
to any remaining uncertainty in measurements of the Galactocentric radius of the
Sun and to variations in the LSR.
